---
title: "A Data Science Approach to Short-Term Energy Demand Forecast"
team: L
session: Hex 5, 2023
coursecode: ZZSC9020
author: 
  - "Louis Long (z5162359)"
  - "Mark Budrewicz (z5353932)"
  - "Hao Chiou (z5131909)"
  - "Charlet Jeyapaul (z5375906)"
date: "07/10/2023"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
## default options of rendering code chunks

# hide underlying code 
knitr::opts_chunk$set(include = F)

# show outputs
knitr::opts_chunk$set(echo = T)

# hide messages and warnings
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
```

```{r code_setup, include=FALSE}
## INSTRUCTIONS

# 1. if need to set working directory, do so below
# setwd(<insert working directory here>)

# 2. open Code/03 Modelling and check the parameters
# in particular, 

# run required scripts
source("Code/01 Setup.R")
source("Code/02 Data exploration.R")
source("Code/03 Modelling.R")
```

\newpage

# Abstract

<Insert abstract here>

# Literature review

<Insert literature review here>

# Material and Methods

## Software

R and Python of course are great software for Data Science. Sometimes, you might want to use `bash` utilities such as `awk` or `sed`.

Of course, to ensure reproducibility, you should use something like `Git` and RMarkdown (or a Jupyter Notebook). Do **not** use Word!

## Description of the Data

How are the data stored? What are the sizes of the data files? How many files? etc.

## Pre-processing Steps

What did you have to do to transform the data so that they become useable?

## Data Cleaning

How did you deal with missing data? etc. 

## Assumptions

What assumptions are you making on the data?

## Modelling Methods

# Exploratory Data Analysis

# Analysis and Results

## Modelling setup

Step 1 - split into training, test and validation sets.

- Show date range of each set
- Explain purpose of each set

```{r modelling_setup}



```

Step 2 - define lagged variables

Step 3 - remove datapoints with missing lagged variables (total removal of one day of data)

## LASSO

Explain scaling variables

Explain use of cross-validation of the combined training and test set

Justify use of nfolds = 5

Show lambda grid considered

Show best lambda selected (lambda vs. MSE chart)

Show shrinkage of coefficients

Show performance on training set

Show example chart of fitted vs. actuals

## Support vector regression (SVR)

Explain scaling variables

Justify features selected

Explain tuning process

Show hyperparameters considered

Show performance of each model and overall model chosen

## Random forest

## Final model selection

# Discussion

Put the results you got in the previous chapter in perspective with respect to the problem studied.

# Conclusion

--

\newpage

# References {-}

<div id="refs"></div>

\bibliographystyle{elsarticle-harv}
\bibliography{references}

# Appendix {-}